Noisy Work: A Review of The Uncertainty Visualisation Literature

### 2024-02-26 paper.qmd
finished word vomit of noise vs signal section

Risk and uncertainty are not only different in how they are communicated, but also why they are communicated. Risk communication focuses on communicating probability around events. These events are typically a binary event that is known to be random, so a version of communication that does not mention risks (e.g. telling someone they will or will not get cancer instead of communicating their risks about it) inherently carries with it a lack of transparency. Additionally risk communication is often communicated to avoid or informa about an unfavorable outcome, so framing can be important if we want to influence decisions in a certainty way. These motivations do not extend to the communication of uncertainty information more broadly. While uncertainty information *is* often used to improve decisions, it seems the main motivation in its communication is to offer transparency to the viewer about a statistical analysis. That is, it acts as a form of statistical "hedging" for signals found in an analysis. Interviews with experts in statistic back up this primary motivation, as ignoring uncertainty information of often expressed as being similar to fraud or lying and the goal of "improving decisions" is only seen as a secondary outcome of an appropriately hedged signal [@Hullman2020a; @Manski2020]. This concept of uncertainty means that the work on communicating risk does not cover the field of uncertainty communication as a whole. It covers the specific case where the signal itself *is* uncertain, but this work does not naturally extend to the goal of hedging a signal (i.e. when we want to communicate uncertainty as noise). Unfortunately understanding the relevant noise seems to be its own problem, as statisticians and visualisation authors seem to be unaware or confused in how to calculate the uncertainty related to an analysis [@Hullman2020a]. Additionally, communicating uncertainty such that it conveys the appropriate singal supression involves combining uncertainty from multiple different sources, something that is considered to be difficult by even experts in uncertainty communication [@Spiegelhalter2017].

@@

Trust is another concept that is related to uncertainty communication, but fact that it is often front and center in the discussion has a somewhat ominous subtext. If the purpose of displaying uncertainty information is to appropriately hedge a signal with noise, then it should be assumed that trust is only related to uncertainty communication through increased transparency and honesty. It does seem to be the case that uncertainty communication increases trust because it is a proxy for *(add citations)*. Unfortunately a large proportion of the literature discusses trust as something that is directly related to uncertainty visualisation rather than as an observable product. In viewing uncertainty communication as directly related to trust, not related through the proxy of transparency, several unobserved variables are conflated. The amount of uncertainty that is present in the information, whether or not the information is trustworthy or makes sense, prior beliefs of the participants, and the trustworthiness of the source are a few examples of variables that are conflated when authors directly consider trust to be of direct interest. Similarly to the issues in risk-aversion directly considering trust, and not transparency, as the metric of importance in uncertainty communication leads to a questionable subtext that argues against transparency. Being concerned about audiences perception of trust, without first establishing if what we are communicating is *trustworthy*, leads to the implication that details that result in information not being considered trustworthy should be avoided.

Authors that imply that a decrease in trust or an increase in ambiguity or risk aversion are metrics of importance in of themselves do not understand the purpose of uncertainty communication as methods to increase transparency. Science communication should be primarily concerned with accuracy, setting trust and risk-aversion as the variables of interest implicitly encourages statisticians to set trust and risk-aversion as the primary goals of communication. The issue of trust being divorced from trustworthiness has been commented on by other authors [@ONeill2018], however the issue still persists in the uncertainty visualisation literature [@Zhao2023]. Additionally, it seems that risk-aversion being divorced from the value of risk has gone unnoticed.

@@

## Uncertainty and Inference
When we consider mapping uncertainty, a reasonable questions that should come to mind is "uncertainty about what?". Every single thing concept or activity in statistics does not have uncertainty. For example, descriptive statistics cannot have uncertainty as they are just describing statistics as they are, the sample mean is the sample mean, we are not uncertain about its value. A sample variance describes dispersion, which is also how we describe uncertainty, but the sample variance in of itself does not have any uncertainty. Most of the mathematical uncertainty concepts have this idea embedded in them. For example, a confidence interval is an interval drawn around the sample mean that will theoretically contain the *populaton mean* 95% of the time. This issue has been noticed several times, however it is often brought up as a *task* dependence rather than an understanding that uncertainty is inherrently linked to inference on a particular statistic. @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Munzner2009 created a nested model for visualisation that highlighted how the first mistake that can be made in a visualisation is in the problem characterisation, and failling to do it well can cause downstream effects and damage the effectivness of a visualisation. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. This should make it clear that uncertainty only exists when we are performing inference. Therefore it only makes sense to describe uncertainty in terms of something to be uncertain *about*, it is not some a latent feature of statistics and data.

@@

- @Zhao2023 displayed a model prediction with uncertainty and took participants using the model prediction as a sign of trust, failing to notice that there is no reason for uncertainty information to influence someone to *use their own prediction*. If the Beuro of Meterology was uncertain about their rain prediction, that would not be sufficient conditions for me to decide to make my own rain predictions. Despite this, the authors seemed to assume that the uncertainty information *should* do that.  

@@

- @`Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included". @Potter2010 aimed to create a summary plot that "concisely presented data with uncertaninty information". These papers fail to realise the exploration step of an analysis, which includes descriptive statistics, exploratory data visualisaiton, and unsupervised machine learning techniques, is performed without a prior hypothesis. Therefore, it is performed to *suggest* a direction for future inference work or communication, which *do* contain uncertainty and does not contain uncertainty within itself. This confusion would not occur if it was understood that uncertainty does not exist without inference.

@@

@@

Uncertainty evaluation studies are not all value extraction. There are cases where authors *know* the uncertainty information is relevant, but are not quite sure how to capture it. These studies most commonly evaluate a visualiation on its ability to convey information for a decision or how well it induces trust.

*decision and trust paragraph*

There are other studies that try to include uncertainty information without directly asking for it. This is an interesting approach, but these papers are often still looking for a "signal" which is specified in the form of a "ground truth" that is used to evaluate the plot. Unfortunately this usually comes in the form of slightly cryptic questions that create a large amount of noise on the interpretation side [@Hullman2016].
- @Hofmann2012 showed two distributions in 20 different visualisations (a lineup protocol) using a jittered sample, a density plot, a histogram, and a box plot. Participants were asked to report in which of the plots was "the blue group furthest to the right" and to provide reasoning from a multiple choice list and grade their certainty. The experiment set up is shown in @fig-right. The participants answers were then compared to a ground truth where the correct plot had a right shifted mean. By comparing the results to a ground truth statistic and marking participants as "wrong" or "right", the error from the participants that had an alternative interpretation to the concept of "furthest right" was conflated with the error from a poor design.

@@

Additionally, the use of a ground truth highlights how participants were supposed to *use* the uncertainty information. In this informa

- The way we ask questions and the types of questions we ask are selected with little justification. This paper makes suggestions to reduce the noise in the data from these papers. @Hullman2016

The view of uncertainty as a dimensionality problem is not quite right. It is important to keep in mind that just because information is in a graphic, that does not mean it will be "seen". The phenomena of inattentional blindness shows that there is no perception without attention and it is powerful enough that participants can fail to be aware of random objects appearing on a screen or a gorilla walking through a basketball game [@simons1999gorillas; @mack2003inattentional]. Therefore if information is mapped to graphical elements that are so low on hierarchy they can be ignored, they might as well not be there at all. The people who don't plot uncertainty because they think its unimportant and the people who relegate uncertainty to the lowest ranking aesthetic on the list are both saying the same thing, "I think the uncertainty is unimportant". Including uncertainty is worth very little if no attention is left to see it. This does not mean we cannot put a *large* amount of information in a graphic. Glyph maps can be used to depict the multi-dimensional information in spatial-temporal data by mapping line plots to map locations, but we still need to decide what information is important. The map can either present trends in the global or local variance depending on whether or not the line plot is scaled globally or locally [@Wickham2012], however smaller details are almost impossible to convey. No matter how much information we try and put in a plot, there will always be only a handful of key takeaways.

- rarely used extension is aggregation of uncertainty over an area (retrieve overall estimation from a spatial distribution of uncertainty). Other tasks include comparisons, rankings. also have a task called "search" where participants have to identify entities that fulfill certainty characteristics (high or low values) @Kinkeldey2014
- Whether to use classed or unclassed uncertainty is rarely discussed @Kinkeldey2014. e.g. land map with remotely sensed images has ambiguity in the boundaries, which can result from multiple sources (land cover class, measurement error, images from different dates). If you combine them all it might not make sense for someone who only needs one.

- depicts a bivariate map which uses a bivariate colour palette that is created by blending two single hue colour palettes. One colour represents the variable of interest while the other represents the sampling error of that variable. There are two immediate problems with this method. First of all, uncertainty is being expressed with hue and saturation which @Maceachren2012 found to be the worst aesthetics to map to uncertainty to as they don't have an intuitive interpretation. Value has a natural connection to uncertainty (lighter values equate to higher uncertainty and darker values equate to more certainty) so it is a much more appropriate choice. While the `Vizumap` data does depict areas of light and darkness, they are largely irrelevant to the uncertainty measure causing our heuristics to lead us to the incorrect conclusions. The Value-Suppressing Uncertainty Palettes (VSUP) shown in @fig-vsup maps estimates to the hue and error to the value thereby creating a more intuitive plot [@Correll2018]. Additionally, at high levels of uncertainty VSUP only has one output colour, which prevents viewers from decrypting any particular value and also avoids enforcing a binary encoding of significance [@Correll2018]. Unfortunately VSUP are not easy to combine with packages like `Vizumap` which leaves it still somewhat difficult to express this encoding in practice, however the combination of a bivariate map with VSUP has shown to improve decisions in the face of uncertainty [@Correll2018].

- @Patrick2023 investigated peoples ability to identify a signal and compared it to typical hypothesis tests

- The line-up protocol is an application on this idea and considers each visualisation to be a single outcome of some larger distribution [@Buja2009; @Wickham2010; @Chowdhury; @Hofmann2012]. By generating a sample of visualisations from a hypothetical distribution, visualisation authors can check if perceived patterns are real or merely the result of chance.

The final consideration when deciding what information to depict is whether or not additional information will clutter the graph. Our visualisation should aim to show enough information to solve a task while avoiding irrelevant distracting information [@kosslyn2006graph]. It also seems that people will discount information they perceive as discountable if they are able. Including mean estimates on depictions of mass can cause people to discount the uncertainty information and use the difference between means as a proxy for the probability distribution [@Kale2021].

This discrepancy is also noted by @Kinkeldey2014 who, in their own literature review on uncertainty methods, comment that the literature seems unsure if uncertainty is simply another variable or if it should be treated differently. This is a fair question to ask. If uncertainty is simply a signal we are trying to convey, why should it be treated any different to any other signal in a visualisation?

- I think the questions where uncertainty is a signal, not a noise, allow the authors of visualisation experiments to ignore the context and motivation that usually governs a visualisation.
- Weirdly When setting their ground truth it seems like a lot of papers EXPECT participants to completely ignore uncertainty information as noise. Like the participants are actually considered to be incorrect if they dampen their signal understanding because of noise considerations which I find FUNNY. (this feels like it is more as a communicating uncertainty issues)
- papers usually have same variance if there is a signal question, and use the mean as a proxy to answer the question. Really should ask a changing signal question where the signal IS impacted by the uncertainty information.
- When asking questions, researchers should focus on events that can be repeated (because of a misunderstanding of confidence intervals) @Hullman2016 (moved here because it is clear she is touching on the whole "what are we uncertain about" thing)

- new papers dont seem to be aware of this problem.
- if the experiment DOES use noise as a signal, it is often very unclear how the participants should include that in their respose and what would be a valid response from participants (e.g. the likelihood to how much question and the cone of uncertainty questions). separate to the data, uncertainty has its own goal. Is it to impact confidence in the estimate (like a bayesian thing)? to supress signal (like a barely noticiable difference line up plot thing)? to facilitate more efficient decisions (so accounting for risk thing)?

- will it impact their decision at all? there are a couple of papers where you can answer all the questions by just ignoring the uncertainty information because of the way they have set it up. e.g. difference in means when the variance is the same, experiments that check if someone can still get the signal and make sure the uncertainty information isnt in the way.

- The literature also does not discuss if uncertainty is another variable, or if it needs to be treated differently (i,.e. is it metadata of another variable). Most papers studies on uncertainty dont consider this issue, and the question remains, should someone have two separate values, or an integrated view uncertain data view @Kinkeldey2014

This is best seen with a simple look at some papers
(Make into table)
Questions where uncertainty is a signal

Questions where uncertainty is a noise

- Should add plot comparison mathematical framework example here
- take a plot and work through the entire example as a flow chart thing
- tasks e.g. "What is the average of X" (how well lined up you are moving down the data pipeline) and resolution (uncertainty is very often moving back a resolution level in statisitcs - also can be an assumption - bootstrap is a replacement for an assumption, not a resolution change, but bootstrapping is centered on the data)  e.g. "What is the noise around the estimate of the value of X"
- "cannot assess the quality of risk communication unless the objectives are clear" circles back to motivation. He assumes we are fulfilling a duty to inform. @Spiegelhalter2017

- heath had a wide range of different metrics they adopted to communicate chronic risk of adverse events in the future @Spiegelhalter2017 (use this for motivation based statistics since they define new statistics based off what they need)
- Mentioned alleatory vs epistemic uncertainty, in the visualisation literature alleatory uncertainty is more commonly the focus. also mentions dependence and joint pdfs as a consideration @Hullman2016
- Most studies involve intrinsic approaches such as colour, transparency, and those that used extrinsic techniques typically used glyphs, error bars, grid based techniques, or contouring @Kinkeldey2014
- most studies used coincident approaches, static visualisations, are applied to a specific domain and dont have generalised results @Kinkeldey2014
- The tasks for participants were most commonly value retrival (some where uncertainty and data value are retrived separately, others where they are some separately). also also be asked to extract a relative value @Kinkeldey2014
- How does the complexity of uncertainty relate to the cateogies of user and task @Kinkeldey2014
- Several studies probide evidence that the usability of uncertainty representations can be highly user and task dependent @Kinkeldey2014
- The nature of the task plays an important role for the usability of uncertainty techniques, this may explain many of the insonsitent outcomes from the studies under review since two studies assessing sifferent techniques and user groups are not comparable @Kinkeldey2014
- Doing a follow up publication to discuss issues with reasoning and decision making based on uncertainty visualisations which is not considered or discussed in this paper @Kinkeldey2014
- Main take away is that "we need to systematize future empirical studies on uncertainty visualisation to better enable comparison and generalisation of the findings"
- One way to advance this goal is a taxonomy, however existing taxonomies are focused on data types, uncertainty categories, and representation types @Kinkeldey2014
- he lists a bunch of tasks that were tested with uncertainty visualisationsat the end @Kinkeldey2014

- risk is uncertainty as a signal, and when uncertainty communication is spoken about this is typically what is discussed, uncertainty as signal supression is what is usually meant when we talk about visualisations but that area is functionally ignored.
This idea, that uncertainty can only be defined in the face of a motivating question, is well grounded in the literature but seldom explicitly mentioned

In section 2.1 it was mentioned that uncertainty must be defined within a specific motivating question, otherwise it inherently does not make sense. A large difficulty with the uncertainty visualisation evalusation studies is that this rule is not followed. There are a shockingly large number of evaluation studies that seem to pay no attention to the information that is relevant to the question they are asking, and show participants a selection of seemingly random visualisations from a statistical point of view.

- inferential uncertainty and outcome uncertainty ARE NOT THE SAME THING??? they visualise DIFFERENT DISTRIBUTIONS

We define two primary motivations for uncertainty visualisation.
  1) To prevent deterministic conclusions from a random signal (uncertainty as noise)
  2) To convey information about a variance, probability, or other random event (uncertainty as signal)

Uncertainty as signal papers have incredibly predictable results
Uncertainty as noise papers should follow a similar protocol to the line up papers

Notes from previous sections that were moved here
- The concept of uncertainty being task dependent is *particularly* salient for uncertainty visualisation, and it is repeatedly identified as a problem in previous reviews of the uncertainty visualisation literature [@Kinkeldey2014; @Hullman2016] as well as across many sub domains and applications [@Wallsten1997; @Munzner2009; @Fischhoff2014; @Meng2021; @Amar2005]. The fact that this conclusion is repeatedly reached shows both the importance and the lack of acknowledgement this concept receives.
- @Fischhoff2014 discusses how decisions with different levels of flexibility should have different uncertainty visualisations. If the user has no choice they should simply be made aware of the risk, if they have several options, the user needs to see the uncertainty in each, if they need to come up with the options themselves, they need to see how things work and how the system fails.
- @Amar2005 suggested a taxonomy for information visualisation based on the types of tasks we use visualisations for and suggest 10 "analytical primitives" that we can then map to visualisations.
- Statistics is, at its core, the study of uncertainty. Therefore discussing uncertainty visualisation a a separate sub domain to "normal" data visualisation is inherently confusing. What is typically meant by "uncertainty" visualisation is "noise", that is, we want to present some signal cushioned by its natural variance. Unfortunately, this distinction between "signal" and "noise" is entirely goal dependent.
- @Bella2005 found that most participants were ignorant to the fact that error bars are used for both confidence intervals and standard error bars, two wildly different indicators of precision. (**also the paper sherry sent me**)

- Need to compare to an approripate ground truth, but this is a philisophical exercse @Hullman2016

>>>>

Risk and uncertainty are not only different in how they are communicated, but also why they are communicated. Risk communication focuses on communicating probability around events. These events are typically a binary event that is known to be random, so a version of communication that does not mention risks (e.g. telling someone they will or will not get cancer instead of communicating their risks about it) inherently carries with it a lack of transparency. Additionally risk communication is often communicated to avoid or information about an unfavorable outcome, so framing can be important if we want to influence decisions in a certainty way. These motivations do not extend to the communication of uncertainty information more broadly. While uncertainty information *is* often used to improve decisions, it seems the main motivation in its communication is to offer transparency to the viewer about a statistical analysis. That is, it acts as a form of statistical "hedging" for signals found in an analysis. Interviews with experts in statistic back up this primary motivation, as ignoring uncertainty information of often expressed as being similar to fraud or lying and the goal of "improving decisions" is only seen as a secondary outcome of an appropriately hedged signal [@Hullman2020a; @Manski2020]. This concept of uncertainty means that the work on communicating risk does not cover the field of uncertainty communication as a whole. It covers the specific case where the signal itself *is* uncertain, but this work does not naturally extend to the goal of hedging a signal (i.e. when we want to communicate uncertainty as noise).

Unfortunately understanding the relevant noise seems to be its own problem, as statisticians and visualisation authors seem to be unaware or confused in how to calculate the uncertainty related to an analysis [@Hullman2020a]. Additionally, communicating uncertainty such that it conveys the appropriate singal supression involves combining uncertainty from multiple different sources, something that is considered to be difficult by even experts in uncertainty communication [@Spiegelhalter2017].

@@

@@

## Uncertainty and Inference
When we consider mapping uncertainty, a reasonable questions that should come to mind is "uncertainty about what?". Every single thing concept or activity in statistics does not have uncertainty. For example, descriptive statistics cannot have uncertainty as they are just describing statistics as they are, the sample mean is the sample mean, we are not uncertain about its value. A sample variance describes dispersion, which is also how we describe uncertainty, but the sample variance in of itself does not have any uncertainty. Most of the mathematical uncertainty concepts have this idea embedded in them. For example, a confidence interval is an interval drawn around the sample mean that will theoretically contain the *populaton mean* 95% of the time. This issue has been noticed several times, however it is often brought up as a *task* dependence rather than an understanding that uncertainty is inherrently linked to inference on a particular statistic. @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Munzner2009 created a nested model for visualisation that highlighted how the first mistake that can be made in a visualisation is in the problem characterisation, and failling to do it well can cause downstream effects and damage the effectivness of a visualisation. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. Health have a wide range of metrics used to communicate different risks [@@Spiegelhalter2017] indicating they utilise new statistics based off the specific aspect of risk that is relevant to the issue. This should make it clear that uncertainty only exists when we are performing inference. Therefore it only makes sense to describe uncertainty in terms of something to be uncertain *about*, it is not some a latent feature of statistics and data.

@@

@@

- @Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included". @Potter2010 aimed to create a summary plot that "concisely presented data with uncertaninty information". These papers fail to realise the exploration step of an analysis, which includes descriptive statistics, exploratory data visualisaiton, and unsupervised machine learning techniques, is performed without a prior hypothesis. Therefore, it is performed to *suggest* a direction for future inference work or communication, which *do* contain uncertainty and does not contain uncertainty within itself. This confusion would not occur if it was understood that uncertainty does not exist without inference.

@@

There is also the common view that uncertainty as a dimensionality problem. This perspective is not quite right. It is important to keep in mind that just because information is in a graphic, that does not mean it will be "seen". The phenomena of inattentional blindness shows that there is no perception without attention and it is powerful enough that participants can fail to be aware of random objects appearing on a screen or a gorilla walking through a basketball game [@simons1999gorillas; @mack2003inattentional]. Therefore if information is mapped to graphical elements that are so low on hierarchy they can be ignored, they might as well not be there at all. The people who don't plot uncertainty because they think its unimportant and the people who relegate uncertainty to the lowest ranking aesthetic on the list are both saying the same thing, "I think the uncertainty is unimportant". Including uncertainty is worth very little if no attention is left to see it. This does not mean we cannot put a *large* amount of information in a graphic. Glyph maps can be used to depict the multi-dimensional information in spatial-temporal data by mapping line plots to map locations, but we still need to decide what information is important. The map can either present trends in the global or local variance depending on whether or not the line plot is scaled globally or locally [@Wickham2012], however smaller details are almost impossible to convey. No matter how much information we try and put in a plot, there will always be only a handful of key takeaways.

This issue has come up in many different forms in the uncertainty visualisation literature. Typically as a "task dependence" on the performance of uncertainty visualisations. It is repeatedly identified as a problem in previous reviews of the uncertainty visualisation literature [@Kinkeldey2014; @Hullman2016] as well as across many sub domains and applications [@Wallsten1997; @Fischhoff2014; @Meng2021; @Amar2005]. The fact that this conclusion is repeatedly reached shows both the importance and the lack of acknowledgement this concept receives. @Fischhoff2014  takes this approach by considerting how decisions with different levels of flexibility should have different uncertainty visualisations. If the user has no choice they should simply be made aware of the risk, if they have several options, the user needs to see the uncertainty in each, if they need to come up with the options themselves, they need to see how things work and how the system fails. @Kinkeldey2014 suggested that the nature of the task plays an important role and believed the task dependence for uncertainty visualisation could explain some of the inconsistency in outcomes.

@@

Uncertainty evaluation studies are not all value extraction. There are cases where authors *know* the uncertainty information is relevant, but are not quite sure how to capture it. These studies most commonly evaluate a visualiation on its ability to convey information for a decision or how well it induces trust. The motivation behind these experiments are obvious, if we cannot measure uncertainty *directy* maybe we can measure its secondary affects. The issue is that in order to say if some plots are *better* or *worse* than others, is to ascribe value to uncertainty information.

Decision making questions that typically involve uncertainty discuss the concept of "risk-aversion" as though it is a mistake. The irony of an uncertainty visualisation pper expressing the importance of uncertainty visualisation, but setting up a ground truth or optimum strategy in which participants *should* ignore the uncertainty information is not lose on us. This perspective does not make sense because it implies participants should make decisions entirely based on the average. A possible soltion for this is to provide a utility framework for a particular experiment [@Hullman2016], however it is unclear how easily participants could adopt a utility framework that is different to their own. @Cheong2016 introduced a very simple optimum strategy into their experiment on bushfire uncertainty visualisations, where participants should choose to evacuate a house as soon as its likelihood of catching fire is 50% or higher. Interestingly, despite this very obvious and simple tactic to maximize payout from the experiment, it seems like many participants did not adopt it, instead acting much more warily as though they were considering whether or not they would *actually* evacuate in the event of a fire. These papers also very rarely consider what the optimum decision would be in a given scenario, and a difficulty with anticipating the best strategy is conflated with the failings of the uncertainty visualisation. Ultimately, we agree with @Hullman2016 who suggested that what determines an approripate ground truth is largely a philosophical exercise.

Authors also occasionally use trust as a proxy for uncertainty. Similar to measuring uncertainty through decision quality, this method also has its own series of issues. If the purpose of displaying uncertainty information is to appropriately hedge a signal with noise, then it should be assumed that trust is only related to uncertainty communication through increased transparency and honesty. It does seem to be the case that uncertainty communication increases trust because it is a proxy for transparency *(add citations)*. Unfortunately a large proportion of the literature discusses trust as something that is directly related to uncertainty visualisation rather than as an observable by product. In viewing uncertainty communication as directly related to trust, not related through the proxy of transparency, several unobserved variables are conflated. The amount of uncertainty that is present in the information, whether or not the information is trustworthy or makes sense, prior beliefs of the participants, and the trustworthiness of the source are a few examples of variables that are conflated when authors directly consider trust to be of direct interest. Considering trust, and not transparency, as the metric of importance in uncertainty communication leads to a questionable subtext that argues against transparency. @Hullman2020a found that author simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as "untrustworthy".Being concerned about audiences perception of trust, without first establishing if what we are communicating is *trustworthy*, leads to the implication that details that result in information not being considered trustworthy should be avoided. Authors that imply that a decrease in trust or an increase in ambiguity or risk aversion are metrics of importance in of themselves do not understand the purpose of uncertainty communication as methods to increase transparency. Science communication should be primarily concerned with accuracy, setting trust and risk-aversion as the variables of interest implicitly encourages statisticians to set trust and risk-aversion as the primary goals of communication. The issue of trust being divorced from trustworthiness has been commented on by other authors [@ONeill2018], however the issue still persists in the uncertainty visualisation literature [@Zhao2023].

Additionally, some studies use a ground truth in these trust experiments, which makes for a very confusing read on exactly how participants are supposed to use the information. 
- @Zhao2023 displayed a model prediction with uncertainty and took participants using the model prediction as a sign of trust. They reported that visualising uncertainty information caused participants to trust the model in the low variance case, but the results in the high variance case were inconclusive. The discussion made it clear the authors thought the uncertainty information should make the visualisation more trustworthy, but conflating trust and the use of a prediction implied uncertaitny information should somehow influence participants to use their own prediction. If the Beuro of Meterology was uncertain about their rain prediction, that would not be sufficient conditions for me to decide to make my own rain predictions. Despite this, the authors seemed to assume that the uncertainty information *should* have an influence on that, showing they had not deeply considered how uncertainty inforormation should infludence the choices of the participants.

A similar measure to trust is using "confidence" in an extracted value or a decision. Interestingly, "confidence" is also used to try and capture the clarity of a message in a normal visualisation. Confidence cannot simultaneously be a measure of clarity of visualisation *and* a way to capture the uncertainty expressed in a visualisation.

There are other studies that try to include uncertainty information without directly asking for it but do not go down the route of asking for a decision or levels of trust. These papers often try to ask a question that should utilise both the uncertainty and signal in the response. This is an interesting approach, but these papers are often still looking for a "signal" which is specified in the form of a "ground truth" that is used to evaluate the plot. Unfortunately this usually comes in the form of slightly cryptic questions that create a large amount of noise on the interpretation side [@Hullman2016].
- @Hofmann2012 showed two distributions in 20 different visualisations (a lineup protocol) using a jittered sample, a density plot, a histogram, and a box plot. Participants were asked to report in which of the plots was "the blue group furthest to the right" and to provide reasoning from a multiple choice list and grade their certainty. The experiment set up is shown in @fig-right. The participants answers were then compared to a ground truth where the correct plot had a right shifted mean. By comparing the results to a ground truth statistic and marking participants as "wrong" or "right", the error from the participants that had an alternative interpretation to the concept of "furthest right" was conflated with the error from a the visualisation choice.
- @Padilla2017 made several confusing assumption when determining the ground truth of their experiments that were assessing storm visualisations. Experiment 2 assumed that the participants should provide responses that indicated they perceived the intensity of the storm to be independent of the uncertainty distribution of its location. This assumption seems counterintuitive as best, however experiment 3 asked participants to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path" for which they *were* supposed to incorperate uncertainty information and not assume they were independent. The authors seemed to be unaware of these confounding mistakes in their assumptions of how the participants were supposed to utilise the uncertainty information.
- @Sanyal2009 mapped uncertainty to dots and signal to a 3D surface and asked participants to identify areas of high and low signal and high and low uncertainty. Participants were not asked to combine that information in any way, and the signal and the noise were treated as separate variables.
- @Correll2014 asked participants to extract the mean and variance from four uncertainty visualisations, bar charts with error bars, box plots, gradient plots, and violin plots (that were adjusted to make sure the mathematical information they were showing was the same) from two side by side distributions. Participants were also asked to answer a question that was supposed to incorperate both the signal and the noise, such as "How likely is candidate B to win the election?" when the two distributions indicated voter preference. Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1=Outcome will be most in favor of A to 7=Outcome will be most in favor of B. The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorperate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.  
- @Cheong2016 tested multiple different visual representations of uncertainty for representing the likelihood of a house being burned down based on its location.

@@

It is unclear how participants are supposed to incorperate uncertainty information into their responses when uncertainty visualisation authors themselves are unsure. Other authors interpret this difficulty as a combination of the uncertainty and task complexity [@Kinkeldey2014], and not as a general confusion. All these examples sever to illustrate that *we dont know how to measure uncertainty* and if we don't know how to measure it, *we cannot evaluate the quality of a visualisation*. It could be argued that a well done uncertainty visualisations should have an imperceptible signal unless the signal would be identified with a hypothesis, but this idea also has a miriad of issues that come with it.  There is already a series of issues with the reject/do not reject concepts in hypothesis testing, so it is already questionable to set the results of a hypothesis test as the ground truth in an uncertainty visualisation experiment, however these may be additional problems that go beyond the binary nature of the issue. @Patrick2023 compared people ability to recognise patterns in a residual plot to typical statistical tests and found human viewers looking at a plot were less sensitive than the typical residual tests. These experiments utilised the lineup protocol which has been suggested as a method to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010; @Chowdhury]. This concept bears similarity to the goal of uncertainty visualisation, except uncertainty visualisation wants to achieve this within a single plot.

The problem of trying to assess uncertainty without knowing *why* we are depiting uncertainty has been noticed by other authors. @Spiegelhalter2017 noted we "cannot assess the quality of risk communication unless the objectives are clear".

# Great Examples
This is not to say all experiments related to uncertainty are unusable. Some work that identifies heuristics are both interesting and particularly related to visualisation of uncertain information. @Newman2012 found that participants were more likely to view points within the bar as more likely than points outside of the bar in bar charts with error bars. Similar effects have been identified in other types of uncertainty displayed. @Padilla2017 found that points that were on an outcome of an ensemble display were perceived as more likely than points not on an outcome, even when the point that was not on a specific outcome of the ensemble was closer to the mean of the uncertainty distribution.

- best things you have seen and why they are the best
- can't have a best practices because you need experimental results and my point is that the experiments aren't testing the right things
- pixel map
  - Similar to what could be good but its computation isnt quite what I would like
  - heaps of sample depictions are animated and that is not always possible, but this is a clear indicator of how you can do a sample visualisaiton
  - hasnt been tested so it is unclear if this actually works at signal supression
- VSUP
  - The purpose of uncertainty is clear in the design and it is signal supression
  - Supression is not the only uncertainty feature we should be interested in (shape is also important which is why I have beef with over smoothing and something visualisation is very good to express) but by being clear in its goal it knows what it is
  - It addresses several key issues with the bivariate map that arrise in general and especially when it is used for visualising uncertainty 
- the census dot map (has been taken down but CNN replicated it [here](https://edition.cnn.com/interactive/2021/us/census-race-ethnicity-map/))
  - I love a good depiction of a sample
  - You can see the sparse regions and the pop 
  - it is similar to the pixel map (could also be singal supression but it is not tested)
  - the sample depiction works well with the interactivity and you can see the signal that is appropriate for a particular level of zoomed in. If you zoom out the points turn into solid colours and show clear signal.
- The new york times class mobility and race animated plot [here](https://www.nytimes.com/interactive/2018/03/19/upshot/race-class-white-and-black-men.html) 
- the animated surface classifications from that satelite images experimental paper
- the climate change scenario uncertainty. There are quite a few of these:
  - [e.g.1](https://earthobservatory.nasa.gov/features/GlobalWarming/page5.php)
  - [e.g.2](https://climatechange.chicago.gov/sites/production/files/2016-07/scenariotempgraph_0.jpg)
  - there are a few examples on the [climate change scenario wiki](https://en.wikipedia.org/wiki/Climate_change_scenario)
  - Good example of uncertainty that has uncertainty from two different sources (human decision in the form of inputs and statistical uncertainty), one of which cant be quantified.
  - a good example in how you can combine two different  types of uncertainty
  - [This](https://www.esri.com/arcgis-blog/products/arcgis-pro/mapping/graphical-cartograms-in-arcgis-pro/) has some good maps and some bonkers ones 
- Exceedance probability map (example where uncertainty that *should* be conveyed as signal is expressed as a signal. It is a typical maths problem that involves)
Notes
- Since a lot of the good visualisations are made by random people creating a new visualisation (likely through back and forth communication which is HOW you get good visualisations) 
- A lot of these show the data instead of an inference which is probably better for this scenario because sample size is important for uncertainty. It is not always appropriate and sometimes dots are not exactly equal to one draw. 
  - comments about smoothing in density plots and bin width in histograms. I do think the sample size should be communicated to people. While it is not the only thing that matters it seems kind of pointless to go to such lengths to hide it.
- Sometimes points are used for things that are NOT samples which can get confusing and honestly I dont like it.

# Future work
This paper has identified several issues in the uncertainty visualisation literature and with those issues, we have several suggestions for future work that would likely be fruitful for the field.
